Running simulations by Alisa

### 2015-10-22 sims/run-sim-Alisa.Rmd
got .rmd to work-ish!

Now to define selection. Here we selection against heterozygotes to be strong (0.5).
```{r set_selection}
	zone_age = c(50)
	qtl=list(chr1=data.frame(traits=c("underdominant"), s = S,pos=c(0.5)))
```
`qtl` is a named list, one element per chromosome,
and each element of the list is a data frame
whose columns give the name of the locus,
the selection coefficient,
and the position in Morgans.

@@

This is a two-step process. First we simulate genotypes:
```{r run_sim_step1}
	sims.s0.1 = lapply(zone_age,sim.zone,n.ind=ninds,n.deme=ndemes,sigma=SIGMA)
```
The output is: a list with hierarchy: [[zone_age]][[c("inds","sp.inds","deme","pars","QTL")]]
	where:
		"inds" is a list of length ninds, such that [[ind]][[chr]][[X1/X2]][[data.frame where col1 = coordinates of rec., col2 = ancestor ID]]
		"sp.inds" is a character vector of length ninds, giving the ancestry given an ancestor ID
		"deme" is an integer vector of length ninds, giving the deme number of an ancestor, given their ancestor ID
		"pars" is an integer that gives the number of generatsion of the simulation
		"QTL" returns the QTL vector used in the simulation

@@

This resulting files can get quite large, but we've saved them here
```{r save_file}
	save(sims.sums,file="~/Projects/HybridZones/ClineProjects/sims_sums_s0.5_sigma1_50*500_tau50.Robj")
```

@@

Now, we get frequencies at each of the sites in the above vector, 
first by getting genotypes at all loci, then using those to calculate an allele frequency for each deme:
```{r get_frequencies}
	genotypes = lapply(sims.sums,function(Z){data.frame(do.call(cbind,lapply(Z$ind.ancest, geno.ind, loci)))})
	freqs = lapply(my.genos,function(X){apply(X,1,function(Z){tapply(Z,cut(1:ncol(X),breaks=seq(0,ncol(X),deme_size)),mean)/2})})

@@

Next, we think about tract lengths in these populations

First we need to decide where the loci under selection are.
Here `qtl` is a named list, one element per chromosome,
and each element of the list is a data frame
whose columns give the name of the locus,
the selection coefficient,
and the position in Morgans.

```{r setup_qtl}
qtl=list(
         chr1=data.frame(
                    traits = c("underdominant"),
                    s = c(0.1),
                    pos = c(0.5)
                )
    )
```

Now, let's set up the population,
which means specifying the number of individuals,
the number of demes,
the width of the zone in real units,
and the dispersal SD ($\sigma$).
Our locus has a selection coefficient of `r qtl[[1]]$s`,
so the cline width will be about `r round(1/sqrt(qtl[[1]]$s),digits=1)`$\sigma$ wide;
so we'll take our zone to be four times that,
and measure distance in units of $\sigma$:
```{r setup_pop}
zone.width <- ceiling(4/sqrt(qtl[[1]]$s))  # in real units
n.ind <- 1000
n.deme <- 50
sigma <- 1 # in real units
deme.spacing <- zone.width/n.deme
deme.locs <- (seq_len(n.deme)-0.5-(n.deme/2))*deme.spacing
sigma.demes <- sigma/deme.spacing # in units of demes
```
The discretized dispersal kernel moves as $\lfloor Z+1/2 \rfloor$,
where $Z \sim N(0,\sigma_\text{demes}^2)$;
let's check this has SD close to $\sigma$:
```{r check_sigma}
zz <- floor( rnorm( 1e6, mean=1/2, sd=sigma.demes ) )*deme.spacing
sd(zz)
```
That's pretty close.

Running a simulation
--------------------

The simulation doesn't know how we're measuring things in real units,
so the sigma we pass needs to be in units of demes.
It returns all the information about the individuals,
but then we use `spBreaks()` to extract which individual is which ancestry.
```{r run_quick_sim}
sim <- sim.zone( n.ind=n.ind, n.deme=n.deme, n.gen=10, sigma=sigma.demes, quiet=TRUE )
sim.sums <- spBreaks(sim)
```

OK, now let's genotype these individuals at some loci:
```{r genotype_sim, fig.width=3*fig.dim, fig.height=1.5*fig.dim}
loci <- seq(0,1,0.05)
loci.cols <- rainbow(12)[cut(abs(loci-0.5),breaks=min((length(loci)-1)/2,8))]
info <- data.frame(
               chr=1,
               bp=floor(loci*1e6),
               pos=loci
           )
genotypes <- geno.pop( sim.sums$ind.ancest, list(chr1=loci) )
```

Now, let's
compute ancestry frequencies,
and plot the resulting clines:
```{r genotype_freqs, fig.width=3*fig.dim, fig.height=1.5*fig.dim}
freqs <- do.call( cbind, tapply( 1:ncol(genotypes), sim$deme, function (kk) {
                    rowMeans(genotypes[,kk])
                } ) )
matplot(deme.locs, t(freqs), type='l', col=loci.cols, xlab="spatial position", ylab="ancestry frequency")
legend("topleft",lty=1,legend=loci,col=loci.cols,title="locus position")
```

We can also look at the local frequencies of the diploid genotypes at each locus:
```{r diploid_freqs}
diploid.freqs <- lapply( seq_along(loci), function (locus) {
                do.call( rbind, tapply( 1:ncol(genotypes), sim$deme, function (kk) {
                    as.vector( tabulate( 1+genotypes[locus,kk], nbins=3 ) )
                } ) )
           } )
# one end
matplot(diploid.freqs[[1]],type='l',main=paste("locus at",loci[1]),lty=1,col=c('red','black','green'))
legend('topright',lty=1,col=c('red','black','green'),legend=c("homozygote","heterozygote","homozygote"))
# the selected locus
sel.locus <- match( qtl$chr1$pos, loci )
matplot(diploid.freqs[[sel.locus]],type='l',main=paste("selected locus at",loci[sel.locus]),lty=1,col=c('red','black','green'))
legend('topright',lty=1,col=c('red','black','green'),legend=c("homozygote","heterozygote","homozygote"))
# the other end
matplot(diploid.freqs[[length(loci)]],type='l',main=paste("locus at",loci[length(loci)]),lty=1,col=c('red','black','green'))
legend('topright',lty=1,col=c('red','black','green'),legend=c("homozygote","heterozygote","homozygote"))
```

>>>>

Now to define selection. Here we selection against heterozygotes to be strong (0.5).
```{r set_selection}
	S  = 0.5
	qtl=list(chr1=data.frame(traits=c("underdominant"), s = S,pos=c(0.5)))
```
`qtl` is a named list, one element per chromosome,
and each element of the list is a data frame
whose columns give the name of the locus,
the selection coefficient,
and the position in Morgans.

@@

This is a two-step process. First we simulate genotypes:
```{r run_sim_step1}
	sims.s0.1 = lapply(zone_age,sim.zone,n.ind=ninds,n.deme=ndemes,sigma=SIGMA)
	#This can take a while, 
```
The output is: a list with hierarchy: [[zone_age]][[c("inds","sp.inds","deme","pars","QTL")]]
	where:
		"inds" is a list of length ninds, such that [[ind]][[chr]][[X1/X2]][[data.frame where col1 = coordinates of rec., col2 = ancestor ID]]
		"sp.inds" is a character vector of length ninds, giving the ancestry given an ancestor ID
		"deme" is an integer vector of length ninds, giving the deme number of an ancestor, given their ancestor ID
		"pars" is an integer that gives the number of generatsion of the simulation
		"QTL" returns the QTL vector used in the simulation

@@

This resulting files can get quite large, but we've saved them here
```{r save_file,cache=TRUE}
	save(sims.s0.1,file="sims_sums_s0.5_sigma1_50by500_tau50.Robj")
```

@@

Now, we get frequencies at each of the sites in the above vector, 
first by getting genotypes at all loci, then using those to calculate an allele frequency for each deme:
```{r get_frequencies}
	genotypes = lapply(sims.sums,function(Z){data.frame(do.call(cbind,lapply(Z$ind.ancest, geno.ind, loci)))})
	freqs = lapply(genotypes,function(X){apply(X,1,function(Z){tapply(Z,cut(1:ncol(X),breaks=seq(0,ncol(X),deme_size)),mean)/2})})
	xx <- (1:ndemes)-0.5-ndemes/2
		matplot(xx,freqs[[1]],col=rainbow(20),lty=1,type="l", main="50 generations",ylab="freq",xlab="deme")
		legend('bottomright',col=rainbow(20),lty=1,legend=loci[[1]],cex=0.75)
		matpoints(-xx,pp,col=rainbow(20),lty=3,type="l")

@@

Next, we think about tract lengths in these populations. The following generates the distribution of lengths 
of continuous ancestry B to the right of the specified locus (`focal_sites`) (in this case the selected locus). 
Note that if the specified locus is of ancestry A, then the length of ancestry B is 0. The first function finds
and measures the chunk in an individual that contains the focal site. The second function generates a 
distribution for each deme.

```{r tract_length_functions}
	focal_sites = lapply(qtl,function(Z){Z$pos})

	get.interval.size = function(IND_DATA=sims.sums[[1]]$ind.ancest[[1]],CHR=1,POS=0.5,ancA = TRUE){
		#return the interval containing focal site for a individual	
		chunk = sapply(IND_DATA[[CHR]],function(X){diff(as.numeric(X[which(X$starts<POS & X$stops>POS),1:2]))})	
		identity = sapply(IND_DATA[[CHR]],function(X){X[which(X$starts<POS & X$stops>POS),3]})
		replace(chunk,which(identity==ancA),0)
	}

	get.deme.chunks = function(IND_DATA=sims.sums[[1]]$ind.ancest, DEME = 1, CHR=1,POS=0.5,ancA = TRUE){
		#get distribution of chunks within a deme
		INDS = IND_DATA[which(sims.s0.1[[1]]$deme==DEME)]
		intervals = do.call(rbind,lapply(INDS,get.interval.size,CHR=CHR,POS=POS,ancA=ancA)) 	
		return(intervals)
	}

```
the output of get.deme.chunks is a nind*2 matrix, where each column is a chromosome. To run:

```{r get_tract_lengths}
	tracts = lapply(1:ndemes,function(X){get.deme.chunks(DEME=X,ancA=F)})

		#function to increase transparency of colors
		add.alpha <- function(col, alpha=1){
		if(missing(col))
		stop("Please provide a vector of colours.")
		apply(sapply(col, col2rgb)/255, 2, 
		function(x) 
		rgb(x[1], x[2], x[3], alpha=alpha)) 
		}

	transparent_rainbow = add.alpha(rainbow(ndemes),0.75)

		D = 1; hist(tracts[[D]], border="black",breaks=seq(0,1,0.05),ylim=c(0,100),xlab="length",main="Distribution of tracts")

		for(D in seq(1,ndemes,5)){
		hist(tracts[[D]], col=transparent_rainbow[D],border=NA,breaks=seq(0,1,0.05),ylim=c(0,50),add=T)	
		}
```
To get LD in these populations, we define `get.genotype` to get the genotypes at two loci in individuals, and `get.LD` to calculate LD in a deme:
```{r LD_functions}
 	get.genotype = function(IND_DATA=sims.sums[[1]]$ind.ancest[[1]],CHR=1,POS1=0.5,POS2=0.5){
	site1 = sapply(IND_DATA[[CHR]],function(X){X[which(X$starts<POS1 & X$stops>POS1),3]})	
	site2 = sapply(IND_DATA[[CHR]],function(X){X[which(X$starts<POS2 & X$stops>POS2),3]})
	return(rbind(site1,site2))
	}

	get.LD = function(IND_DATA= sims.sums[[1]]$ind.ancest,DEME=1,CHR=1,POS1 = 0.5, POS2=0.1){
	INDS = IND_DATA[which(sims.s0.1[[1]]$deme==DEME)]
	genotypes = do.call(cbind,lapply(INDS,get.genotype,CHR=CHR,POS1=POS1,POS2=POS2))
	freqB1 = length(which(genotypes[1,]))/length(genotypes[1,])
	freqB2 = length(which(genotypes[2,]))/length(genotypes[2,])
	freqB12 = length(which(genotypes[1,] & genotypes[2,]))/length(genotypes[1,])
	LD = freqB12 - freqB2*freqB1
	return(LD)
}

```
We can just calculate LD over demes and loci pairs. Here's an example of LD between pairs of loci where one member of the
pair is selected on. This gives a plot of decay in population LD as we get further away from the center of the zone.

```
LD_matrix = do.call(cbind, lapply(seq(0.51,0.99,0.05),function(Z){sapply(1:ndemes,get.LD,POS1=0.5,POS2=Z,CHR=1,IND_DATA=sims.sums[[1]]$ind.ancest)}))
	matplot(xx,LD_matrix,type="l",col=rainbow(20),xlim=c(-10,10),lty=1)

Simulate a realistic population (with Populus as a model)
-------------------------
We propose that the density of Populus along a one-dimensional transect is one every 10 meters. Additionally, based on a
literature search that suggests most pollen only makes it to the nearest neighbor and seeds do not disperse much, with 
potentially a long tail, we propose a sigma of 100m. Furthermore, based on HZ width of 20km for most of the geome, we 
propose that width at selected loci may be 1km, giving a selection coefficient of 0.01 (based on width = 2*sigma/sqrt(S)). 
We space demes ever 1km for 100km, with 100 individuals per deme (given above density of 1/10 inds/meter). In units of demes, sigma is 10.

```{r set_populus_parameters,cache=TRUE}
	zone_age	= 10
	ndemes 		= 100
	deme_size 	= 100
	ninds 		= ndemes*deme_size
	SIGMA 		= 0.1
	S 			= 0.01
	qtl=list(chr1=data.frame(traits=c("underdominant"), s = S,pos=c(0.5)))

	sims.populus_like.10gens = lapply(zone_age,sim.zone,n.ind=ninds,n.deme=ndemes,sigma=SIGMA)
	sims.sums.populus_like.10gens = lapply(sims.populus_like.10gens,spBreaks)
	save(sims.populus_like.10gens,file="simspopulus_like.10gens.Robj")
```
